Learning a reward function from demonstrations: a cascaded supervised learning approach

### 2012-10-30 piot2012cascading/ECML13_cascading.tex
Grands changements paris

% Created 2012-10-15 lun. 14:49
\documentclass[smallextended]{svjour3}
\smartqed 
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage[integrals]{wasysym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subfigure}
\tolerance=1000
\usepackage{amsmath}
%\usepackage{amsthm}
%\newtheorem{definition}{Definition}
%\newtheorem{theorem}{Theorem}
\usepackage{dsfont}
\providecommand{\alert}[1]{\textbf{#1}}
\newcommand{\E}{\mathbf{E}}
\newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
\newcommand{\Card}{\operatorname*{Card}} %\operatorname* pour les op. pouvant admettre des limites...
\begin{document}
\title{Learning a reward function from demonstrations: a cascaded supervised learning approach}
\author{Edouard Klein$^{1,2}$ \and Bilal Piot $^{2,3}$\and Matthieu Geist $^{2}$\and Olivier Pietquin$^{2,3}$}
\titlerunning{Cascading approach to IRL}
\institute{
 1. ABC Team\\
 LORIA-CNRS, France.\\
\and
2.Supélec\\
 IMS-MaLIS Research group, France\\
 \texttt{firstname.lastname@supelec.fr}\\
\and 
3. UMI 2958 GeorgiaTech-CNRS\\
France\\
}
\date{\today}

@@

\begin{abstract}
  This paper considers the Inverse Reinforcement Learning problem, that is inferring a reward function for which a demonstrated expert policy is optimal.
We propose to break the IRL problem down into two generic Supervised Learning steps: this is the cascaded approach. A classification step followed by a regression step output a reward function.
This approach is justified by an analysis which shows that the demonstrated expert policy is near-optimal for the output reward function. We explain some advantages of this approach over existing algorithms, mainly not needing to repeatedly solve the Reinforcement Learning direct problem. This makes the proposed approach very competitive to the existing algorithms.
Finally, up to the use of some heuristics, we empirically show the algorithm to work with only transitions sampled according to the demonstrated expert policy. We quantitavely demonstrate the soundness of this approach on a highway driving simulator, and illustrate some aspects of the proposed algorithm on the classical inverted pendulum benchmark.
  \end{abstract}
\section{Introduction}
\label{sec-2}

Inverse Reinforcement Learning (IRL)~\cite{russell1998learning} aims at inferring a reward function for which a demonstrated expert policy is optimal. An important part of the literature (see Sec.~\ref{sec:related} for a brief overview) tries to find a reward function such that the associated optimal policy induces a distribution over trajectories (or some measure of this distribution, typically, the feature expectation) starting in a certain state space\footnote{Formally, there is a distribution over the starting state.} which matches the one induced by the expert. Most of these algorithms use Reinforcement
Learning (RL) algorithms as subroutines as finding the optimal policy for a reward function is a necessary step before computing a measure of the previously mentionned distribution. Thus, their performance depends strongly on the quality of these subroutines. Consequently, they suffer from the same challenges of large state spaces, lack of data etc. as RL. In order to avoid repeatedly solving RL problems, we decide to take a different point of view.

Having in mind that there is a one to one relation between a reward function and its associated optimal action-value function (via the inverse of the Bellman equation, see Eq.~\ref{eq:bellman1}), it is worth thinking of a method which is able to output an action-value function for which the greedy policy is the demonstrated expert policy. Thus, the demonstrated expert policy will be optimal for the reward function associated to the action-value function. As a method which is able to output an action-value function for which the greedy policy is the demonstrated expert policy, we propose a score function-based multi-class classification step (see Sec.~\ref{sec:algo}). Besides, in order to retrieve, via the inverse of the Bellman optimality equation, the reward associated to the score function the classification outputs, we introduce a regression step (see Sec.~\ref{sec:algo}). That is why the method is called the Cascaded Supervised Approach (CSA). This method is justified by a theoretical analysis (see Sec.~\ref{sec:analysis}) where it is shown that the demonstrated expert policy is near-optimal for the reward the regression step outputs.

Finally, an instantiation of CSA is proposed Sec.~\ref{sec:experiments}. This algorithm does not need to iteratively solve a RL problem and needs only sampled transitions from expert and non-experts policies as inputs. Moreover, up to the use of some heuristics, the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy. The algorithm is tested on a car driving simulator and compared to an apprenticeship learning algorithm \cite{abbeel2004apprenticeship}, to a pure classification algorithm \cite{taskar2005learning} and to a recent successfull IRL method \cite{klein2012scirl}.

\section{Background and notations}
\label{sec:background}
%% \begin{itemize}
%% \item Parler du domaine de la prise de décision séquentielle
%% \item Expliquer le problème direct, en profiter pour introduire l'équation d'évaluation de Bellman et l'équation d'optimalité de Bellman
%% \begin{itemize}
%% \item $RL(MDP=\{S,A,\gamma,(P),R\}) = \pi^*:S\times A \rightarrow \mathbb{R}$
%% \item $\forall \pi, Q^\pi(s,a) = R(s,a) + \sum_{s'}P(s'|s,a)Q^\pi(s',\pi(s'))$
%% \item $\pi^* = \arg\max_aQ^*(s,a)$
%% \end{itemize}
%% \item Introduire le problème inverse dans une forme mathématique relativement propre
%% \begin{itemize}
%% \item $IRL( MDP\backslash R = \{S,A,\gamma,(P)\} \cup \{\pi^E\} = R : S\times A \rightarrow \mathbb{R}$ telle que $\pi^E$ est optimale pour $R$
%% \item Mentionner la nature mal posée du problème sans s'appesentir dessus. Préciser toutefois qu'aucune approche ne le résoud de manière mathématiquement satisfaisante à cause de ça.
%% \item insister sur l'optimalité de l'expert comme hypothèse. Un expert sous optimal, c'est un autre champ de littérature, éventuellement donner deux trois refs.
%% \end{itemize}
%% \end{itemize}
First, we introduce some general notations.
Let $E$ and $F$ be two non-empty sets, $E^F$ is the set of functions from $F$ to $E$.
We note $\Delta_X$ the set of distributions over $X$.
Let $\alpha\in\mathbb{R}^X$ and $\beta\in\mathbb{R}^X$: $\alpha\leq\beta \Leftrightarrow \forall x\in X, \alpha(x) \leq \beta(x)$. We will often slightly abuse the notations and consider (where applicable) most objects as if they were matrices and vectors indexed by the set they operate upon.

We work with finite \emph{Markov Decision Processes} (MDPs) \cite{puterman1994markov}, that is tuples $\{S,A,P,R,\gamma\}$. The state space is noted $S$, $A$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in ]0,1[$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the Markovian dynamic of the MDP. Thus for each $(s,a)\in S\times A$, $P(.|s,a)$ is a distribution over $S$ and $P(s'|s,a)$ is the probability to reach $s'$ by choosing the action $a$ in the state $s$. At each time step $t$, the agent uses the information encoded in the state $s_t\in S$ in order to choose an action $a_t \in A$ according to a (deterministic\footnote{We restrict ourselves here to deterministic policies, but the loss of generality is minimal as there exists at least one optimal deterministic policy.}) policy $\pi\in A^S$. The agent then steps to a new state $s_{t+1}\in S$ according to the Markovian transition probabilities $P(s_{t+1}|s_t,a_t)$. The reward function $R$ is a local measure of the quality of the control. The global quality of the control induced by a policy $\pi$, with respect to a reward $R$ is assessed by the value function $V^\pi_R \in \mathbb{R}^{S}$ which associates each state with the discounted cumulative reward for following policy $\pi$ from this state: $V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t,\pi(s_t))|s_0 = s]$. Therefore, an optimal policy $\pi^*_R$ is a policy whose value function (the optimal value function $V^*_R$) is greater than that of any other policy, for all states : $\forall \pi, V^*_R\geq V^\pi_R$. The stationary distribution induced by a policy $\pi$ is written $\rho_\pi,\textrm{ s.t. }\rho_\pi^TP^\pi = \rho_\pi^T$.

The Bellman evaluation operator $T^\pi_R: \mathbb{R}^{S} \rightarrow  \mathbb{R}^{S}$ is defined by
\begin{equation}
  T^{\pi}_RV = R_\pi + \gamma P_\pi V
  \end{equation}
where $P_\pi$ is $(P(s'|s,\pi(s)))_{s,s' \in S}$ and $R_\pi$ is $(R(s,\pi(s)))_{s\in S}$. The Bellman optimality operator follows naturally : $T^*_RV = \max_\pi T^\pi_RV$. Both operators are contractions. The fixed point of the Bellman evaluation operator $T^\pi_R$ is the value function of $\pi$ with respect to reward R: $V^\pi_R$, the fixed point of the Bellman optimality operator $T^*_R$ is the optimal value function $V_R^*$ with respect to reward $R$. Another object object of interest is the action-value function $Q^\pi_R\in\mathbb{R}^{S\times A}$ that adds a degree of freedom on the choice of the first action, formally defined by $Q^\pi_R(s,a) = T^a_RV^\pi_R(s)$, with $a$ the policy that always returns action $a$ ($T^a_RV = R_a + \gamma P_a V$ with $P_a = (P(s'|s,a))_{s,s' \in S}$) and $R_a$ = $(R(s,a))_{s\in S}$. Once expanded, this gives what we refer to as the Bellman optimality equation :
\begin{equation}
  Q^*_R = R_a + \gamma P_{\pi^*_R}V^*_R
  \label{eq:bellman1}
\end{equation}

An optimal policy follows a greedy mechanism with respect to its optimal action-value function 
\begin{equation}
  \label{eq:greedy}
  \pi^*_R(s)\in\argmax_aQ^*_R(s,a).
\end{equation}

When the state space is too large to allow matricial representation or when the transition probability or even the reward function are unknown except through observations gained by interacting with the system, RL or approximate dynamic programming may be used to approximate the optimal control policy~\cite{sutton1998reinforcement}. We call this the direct problem.

In this work, we wish to tackle the inverse problem. We observe an expert's deterministic\footnotemark[\value{footnote}] policy $\pi^E$, assuming that there exist some unknown reward $R^E$ for which the expert is optimal. The suboptimality of the expert is an interesting setting, that has been discussed for example in \cite{melo2010analysis,syed2012reduction}, but that we are not addressing here due to the numerous complications it introduces. We do not try to find this unknown reward $R^E$ but rather a non trivial reward $R$ for which the expert is at least near-optimal. The trivial reward $0$ is a solution to this ill-posed problem (no reward means that every behavior is optimal). Because of its ill-posed nature, this expression of \emph{Inverse Reinforcement Learning} (IRL) still has to find a satisfactory solution although lots of progress have been made, see Sec.~\ref{sec:related}.
\section{The cascading algorithm}
\label{sec:algo}
%% \begin{itemize}
%% \item Mentionner (sans en parler, ce sera fait dans \hyperref[sec-7-1]{Imitation}) l'imitation. Une approche supervisée évidente est la classification, avec pour entrée un set $D_C$ de données issues de l'expert. Rappeler (expliqué pour la première fois dans \hyperref[sec-3]{Formalisme} qu'on se restreint (sans perte de généralité, le préciser) au cas des politique déterministes.
%% \item Tous les classifieurs utilisent une fonction de score (je n'ai pas trouvé un seul exemple d'un classifieur qui n'en utilise pas), il s'agit juste d'un manière d'écrire les choses.
%% \item Expliquer que l'on identifie le mécanisme de classification au mécanisme glouton qui régit la politique optimale (donc celle de l'expert).
%% \item Sans chercher à justifier mathématiquement la chose (ce sera fait dans la partie \hyperref[sec-5]{Analyse}), expliquer qu'intuitivement l'on va identifier fonction de score du classifieur et fonction de qualité de l'expert (qui est optimale pour une certaine récompense, rappelons-le).
%% \begin{itemize}
%% \item Repartir de l'équation d'évaluation de Bellman pour isoler le terme $R(s,a)$ et introduire $r:S\times A\in \mathbb{R}$, la récompense issue de cette équation lorsque l'on remplace $Q^E$ par la fonction de score : $r(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi(s')) $
%% \item On estime ce terme à partir des transitions $D_R$ car les probabilités de transitions sont inconnues : $\hat r(s,a) = q(s,a)-\gamma \sum_{s'} {1\over N_{s'}}\sum_{(s,a,s',a')\in D_R}q(s',a')$
%% \item La fonction de récompense est donnée par une régression sur $D_R \cup {\hat r}$.
%% \end{itemize}
%% \item Donner l'algorithme l'environnement \texttt{algorithm2e}
%% \item Rapidement, l'étape de régression permet d'ajouter la structure du MDP au résultat de classif, en utilisant l'équation de Bellman.
%% \item Bien préciser qu'il ne s'agit ici que d'une explication qualitative, ``avec les mains'', la preuve du fait que ça résoud bien le problème est donnée dans \hyperref[sec-5]{Analyse}, une démonstration sur des problèmes si complexes que l'état de l'art ne peut les résoudre (sauf SCIRL) est donnée dans \hyperref[Experience]{Experience}. Les avantages comparés sont expliqués dans \hyperref[Autres-Approches]{Autres Approches}. Enfin, des explications plus intuitives sur les mécanismes internes de l'algorithme, ainsi que des résultats expérimentaux supplémentaires sont fournis en \hyperref[sec-9]{Annexes}.
%% \end{itemize}
Our first step in the quest for a reward function solving the IRL problem is a classification step using a \emph{score function-based multi-class classifier} (MC$^2$). This classifier learns a score function $q\in\mathbb{R}^{S\times A}$ that rates the association of a given label $a\in A$ with a certain input $s\in S$. The classification rule $\pi_C\in A^S$ is simply to select (one of) the label(s) that achieves the highest score for the given inputs :
\begin{equation}
  \label{eq:greedy2}
\pi_C(s) \in \argmax_a q(s,a).
\end{equation}
For example, \emph{Multi-class Support Vector Machines}~\cite{guermeur2011generic} can be seen as MC$^2$ algorithms, we however used a structured margin approach. We give more examples and detail our choice of a MC$^2$ algorithms for the experiments in Subsec.~\ref{subsec:instanciation}.

Given a dataset $D_C = \{(s_i,a_i=\pi^E(s_i))_i\}$ of actions $a_i$ (deterministically) chosen by the expert on states $s_i$, we train such a classifier. The classification policy $\pi_C$ is not the end product we are looking for (that would be mere imitation of the expert, not IRL). What is of particular interest to us is the score function $q$ itself. One can easily notice the similarity between Eq.~\ref{eq:greedy2} and Eq.~\ref{eq:greedy} that describes the relation between the optimal policy in a MDP and its optimal action-value function. As the expert is optimal with respect to some unknown reward function $R_E$ and we seek a reward function $R$ for which it is also optimal, it is only natural to view the score function $q$ of the classifier as some kind of optimal action-value function for the classifier policy $\pi_C$ (which can itself be seen as a good approximation of $\pi_E$). By inversing the Bellman equation, one can express the reward as a function of the action-value function : $R(s,a) = Q^*_R(s,a) - \sum_{s'}P(s'|s,a)Q^*_R(s'\pi^*(s'))$. This leave us with $R^C$, the reward function relative to our score/action-value function $q$ :
\begin{equation}
  R^C(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi_C(s')).
\end{equation}
We wish to solve the general, or approximate, IRL problem where the transition probabilities $P$ are unknown. For this we use another dataset $D_R = \{(s_j,a_j,s'_j)_j\}$ where $s'_j$ is the state an agent taking action $a_j$ in state $s_j$ transitioned to. Action $a_j$ need not be chosen by any particular policy. The dataset $D_R$ brings us information about the dynamics of the system. Our reward function $R^C$ must be approximated with the help of the information in $D_R$. This is done via datapoints $\{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j$ that we then feed to a regressor (a simple least-square approximator can do) to obtain $\hat R^C$, a generalisation of $\{(\hat r_j)_j\}$ over the whole state-action space. The complete algorithm is given Alg.~\ref{algo:cascading}.
\begin{algorithm}%[H]
    %\small
  \caption{CSA-IRL algorithm}
  \label{algo:cascading}
  \emph{\textbf{Given}} a training set $D_C=\{(s_i,a_i=\pi^E(s_i))\}_{1\leq i \leq D}$ and another training set $D_R=\{(s_{j},a_{j},s'_{j})\}_{1\leq j \leq D'}$\;\\
  \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$\;\\
  \emph{\textbf{Learn}} a reward function $\hat R^C$ from the dataset $\{(s_{j},a_{j},\hat{r}_j)\}_{1\leq j \leq D'}$, $\forall (s_j,a_j,s'_j) \in D_R,\hat{r}_j=q(s_{j},a_{j})-\gamma q(s'_{j},\pi_C(s'_{j}))$\;\\
  \emph{\textbf{Output}} the reward function $\hat R^{C}$ \;
\end{algorithm}

Cascading two supervised approaches like we do is a way to inject the MDP structure in the resolution of the problem. Indeed, mere classification only takes into account information from the expert (i.e., which actions goes with which state), but there is more to a MDP than a single policy. Using the Bellman equation in the expression of $\hat r$, we make use of the information lying in the transitions $(s,a,s')$, namely information about the transition probabilities $P$. The final regression step is just a way to try to generalize this information about $P$ to the whole state-action space, in order to have a well-behaved reward function.

As the analysis (Sec.~\ref{sec:analysis}) reveals, both classification error and regression error should be controlled. By carefully choosing or tuning both supervised learning algorithms, it should be possible to mitigate any ill effects of the specific settings the cascading algorithm is used in. If one operates in a well-known domain where a lot of data is available, then ease-of-use will probably dictate the choice of two off-the-shelf implementations with default settings. If data is lacking, one could for example have to overcome bias, variance or generalization problems in the regression step (more about that in Subsec.\ref{subsec:instanciation}). The possibility to use and leverage the wide range of techniques developped for these two well-known supervised problems is a strong advantage of the proposed cascading approach.

The explanation in this section is only a quatitative attempt to make the reader understand the idea behind the algorithm. The analysis of the next section tackles things in a formal, yet less intuitive, way.

>>>>

% Created 2012-10-15 lun. 14:49
\documentclass[smallextended]{svjour3}
\smartqed 
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{soul}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage[integrals]{wasysym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subfigure}
\tolerance=1000
\usepackage{amsmath}
%\usepackage{amsthm}
%\newtheorem{definition}{Definition}
%\newtheorem{theorem}{Theorem}
\newtheorem{cor}{Corollary}
\usepackage{dsfont}
\usepackage{mathtools}
\mathtoolsset{showonlyrefs=true}
\providecommand{\alert}[1]{\textbf{#1}}
\newcommand{\E}{\mathbf{E}}
\newcommand{\argmax}{\operatorname*{argmax}} %\operatorname* pour les op. pouvant admettre des limites...
\newcommand{\Card}{\operatorname*{Card}} %\operatorname* pour les op. pouvant admettre des limites...
\begin{document}
\title{Learning a reward function from demonstrations: a cascaded supervised learning approach}
\author{Edouard Klein$^{1,2}$ \and Bilal Piot $^{2,3}$\and Matthieu Geist $^{2}$\and Olivier Pietquin$^{2,3}$}
\titlerunning{Cascading approach to IRL}
\institute{
 1. ABC Team\\
 LORIA-CNRS, France.\\
\and
2.Supélec\\
 IMS-MaLIS Research group, France\\
 \texttt{firstname.lastname@supelec.fr}\\
\and 
3. UMI 2958 GeorgiaTech-CNRS\\
France\\
}
\date{\today}

@@

\begin{abstract}
  This paper considers the Inverse Reinforcement Learning problem, that is inferring a reward function for which a demonstrated expert policy is optimal.
We propose to break the IRL problem down into two generic Supervised Learning steps: this is the Cascaded Supervised IRL (CSI) approach. A classification step that defines a score function is followed by a regression step that outputs a reward function.
An analysis of this approach shows that the demonstrated expert policy is near-optimal for the computed reward function.
Not needing to repeatedly solve a Markov Decision Process (MDP) and the ability to leverage existing techniques for classification and regression are two important advantages of the CSI approach. It is furthermore empirically demonstrated to work with only transitions sampled according to the expert policy, up to the use of some heuristics. For this purpose two classical benchmarks (a highway driving simulator and a inverted pendulum simulator) are used.
  \end{abstract}
\section{Introduction}
\label{sec-2}
The sequential decision making problem consists in taking the appropriate action pondering the available data in order to maximize a certain criterion. When framed in an MDP (see Sec.~\ref{sec:background}), the criterion to maximize is the expected sum of discounted rewards. The Inverse Reinforcement Learning (IRL)~\cite{russell1998learning} problem, which we tackle here, aims at inferring a reward function for which a demonstrated expert policy is optimal.

IRL is often used as a way to do Apprenticeshipe Learning (AL): imitating a demonstrated expert policy, without explicitely looking for the reward function. The reward function nevertheless is of interest in and of itself. As mentionned in \cite{ng2000algorithms}, its semantics can be analyzed in biology or econometrics. Practically, the reward can be seen as a succinct description of a task. Discovering it removes the coupling between understanding the task and learning how to fulfill it that exists in AL. IRL allows the use of (Approximate) Dynamic programming ((A)DP) or Reinforcement Learning (RL) techniques to learn the task from the computed reward function. Although some approaches use it as a byproduct, most AL algorithm make not use of the reward function. A very straightforward way to do AL is to use a multi-class classifier to directly learn the expert policy. We compare this idea with our IRL algorithm in the experiments (Sec.~\ref{sec:experiments}).

A lot of existing approaches in either IRL or AL need to repeatedly solve the underlying MDP. Thus, their performance depends strongly on the quality of the associated subroutine. Consequently, they suffer from the same challenges of scalability, data scarcity, etc., as RL and (A)DP. In order to avoid repeatedly solving such problems, we adopt a different point of view.

Having in mind that there is a one to one relation between a reward function and its associated optimal action-value function (via the Bellman equation, see Eq.~\ref{eq:bellman1}), it is worth thinking of a method able to output an action-value function for which the greedy policy is the demonstrated expert policy. Thus, the demonstrated expert policy will be optimal for the related reward function. We propose a score function-based multi-class classification step (see Sec.~\ref{sec:algo}) to infer a score function. Besides, in order to retrieve via the Bellman equation the reward associated to the score function computed by the classification step, we introduce a regression step (see Sec.~\ref{sec:algo}). That is why the method is called the Cascaded Supervised Inverse reinforcement learning (CSI). This method is analyzed, see Sec.~\ref{sec:analysis} where it is shown that the demonstrated expert policy is near-optimal for the reward the regression step outputs.

This algorithm does not need to iteratively solve an MDP and needs only sampled transitions from expert and non-experts policies as inputs. Moreover, up to the use of some heuristics (see Subsubsec.~\ref{subsubsec:heuristics}), the algorithm is able to be trained only with transitions sampled from the demonstrated expert policy. A specific instantiation of CSI (proposed Subsec.~\ref{subsec:instanciation}) is tested on an inverted pendulum simulator (Subsec~\ref{subsec:IP}) and a car driving simulator (Subsec~\ref{subsec:Highway}) where we compare it with an AL algorithm \cite{abbeel2004apprenticeship}, with a pure classification algorithm \cite{taskar2005learning} and with a recent successfull IRL method \cite{klein2012scirl} as well as with a random baseline.

Differences and similarities with existing AL or IRL approaches are succinctly exposed in Sec.~\ref{sec:related}.
\section{Background and notations}
\label{sec:background}
First, we introduce some general notations.
Let $E$ and $F$ be two non-empty sets, $E^F$ is the set of functions from $F$ to $E$.
We note $\Delta_X$ the set of distributions over $X$.
Let $\alpha\in\mathbb{R}^X$ and $\beta\in\mathbb{R}^X$: $\alpha\geq\beta \Leftrightarrow \forall x\in X, \alpha(x) \geq \beta(x)$. We will often slightly abuse the notations and consider (where applicable) most objects as if they were matrices and vectors indexed by the set they operate upon.

We work with finite MDPs \cite{puterman1994markov}, that is tuples $\{S,A,P,R,\gamma\}$. The state space is noted $S$, $A$ is a finite action space, $R\in\mathbb{R}^{S\times A}$ is a reward-function, $\gamma\in ]0,1[$ is a discount factor and $P\in \Delta_{S}^{S\times A}$ is the Markovian dynamic of the MDP. Thus for each $(s,a)\in S\times A$, $P(.|s,a)$ is a distribution over $S$ and $P(s'|s,a)$ is the probability to reach $s'$ by choosing the action $a$ in the state $s$. At each time step $t$, the agent uses the information encoded in the state $s_t\in S$ in order to choose an action $a_t \in A$ according to a (deterministic\footnote{We restrict ourselves here to deterministic policies, but the loss of generality is minimal as there exists at least one optimal deterministic policy.}) policy $\pi\in A^S$. The agent then steps to a new state $s_{t+1}\in S$ according to the Markovian transition probabilities $P(s_{t+1}|s_t,a_t)$. The stationary distribution over the states $\rho_\pi$ induced by a policy $\pi$ satisfies
\begin{equation}
  \rho_\pi^TP_\pi = \rho_\pi^T,
  \end{equation}
where $P_\pi$ is $(P(s'|s,\pi(s)))_{s,s' \in S}$. The stationary distribution relative to the exert policy $\pi_E$ is $\rho_E$.

The reward function $R$ is a local measure of the quality of the control. The global quality of the control induced by a policy $\pi$, with respect to a reward $R$, is assessed by the value function $V^\pi_R \in \mathbb{R}^{S}$ which associates to each state the discounted cumulative reward for following policy $\pi$ from this state:
\begin{equation}
V^\pi_R(s) = \E[\sum_{t\geq 0}\gamma^tR(s_t,\pi(s_t))|s_0 = s].
\end{equation}
This long-term criterion is what is being optimized when solving an MDP. Therefore, an optimal policy $\pi^*_R$ is a policy whose value function (the optimal value function $V^*_R$) is greater than that of any other policy, for all states : $\forall \pi, V^*_R\geq V^\pi_R$..

The Bellman evaluation operator $T^\pi_R: \mathbb{R}^{S} \rightarrow  \mathbb{R}^{S}$ is defined by
\begin{equation}
  T^{\pi}_RV = R_\pi + \gamma P_\pi V
  \end{equation}
where $R_\pi$ is $(R(s,\pi(s)))_{s\in S}$. The Bellman optimality operator follows naturally
\begin{equation}
  T^*_RV = \max_\pi T^\pi_RV.
\end{equation}
  Both operators are contractions. The fixed point of the Bellman evaluation operator $T^\pi_R$ is the value function of $\pi$ with respect to reward R :
  \begin{eqnarray}
    V^\pi_R &=& T^\pi_R V^\pi_R\\
    V^\pi_R &=& R_\pi + \gamma P_\pi V^\pi_R.
  \end{eqnarray}
The Bellman optimality operator $T^*_R$ also admits a fixed point, the optimal value function $V_R^*$ with respect to reward $R$.

Another object of interest is the action-value function $Q^\pi_R\in\mathbb{R}^{S\times A}$ that adds a degree of freedom on the choice of the first action, formally defined by $Q^\pi_R(s,a) = T^a_RV^\pi_R(s)$, with $a$ the policy that always returns action $a$ ($T^a_RV = R_a + \gamma P_a V$ with $P_a = (P(s'|s,a))_{s,s' \in S}$ and $R_a$ = $(R(s,a))_{s\in S}$). The value function $V^\pi_R$ and the action-value function $Q^\pi_R$ are quite directly related : $\forall s \in S, V^\pi_R(s) = Q^\pi_R(s,\pi(s))$. The Bellman evaluation equation for $Q^\pi_R$ is therefore :
\begin{equation}
  Q^\pi_R(s,a) = R(s,a) + \gamma \sum_{s'\in S}P(s'|s,a) Q(s',\pi(s')).
  \label{eq:bellman1}
\end{equation}

An optimal policy follows a greedy mechanism with respect to its optimal action-value function $Q^*_R$
\begin{equation}
  \label{eq:greedy}
  \pi^*_R(s)\in\argmax_aQ^*_R(s,a).
\end{equation}

When the state space is too large to allow matricial representation or when the transition probabilites or even the reward function are unknown except through observations gained by interacting with the system, RL or ADP may be used to approximate the optimal control policy~\cite{sutton1998reinforcement}.

We recall that solving the MDP is the direct problem. This contribution aims at solving the inverse one. We observe an expert's deterministic\footnotemark[\value{footnote}] policy $\pi_E$, assuming that there exists some unknown reward $R_E$ for which the expert is optimal. The suboptimality of the expert is an interesting setting that has been discussed for example in \cite{melo2010analysis,syed2010reduction}, but that we are not addressing here. We do not try to find this unknown reward $R_E$ but rather a non trivial reward $R$ for which the expert is at least near-optimal. The trivial reward $0$ is a solution to this ill-posed problem (no reward means that every behavior is optimal). Because of its ill-posed nature, this expression of \emph{Inverse Reinforcement Learning} (IRL) still has to find a satisfactory solution although lots of progress have been made, see Sec.~\ref{sec:related}.
\section{The cascading algorithm}
\label{sec:algo}
Our first step towards a reward function solving the IRL problem is a classification step using a \emph{score function-based multi-class classifier} (SFMC$^2$ for short). This classifier learns a score function $q\in\mathbb{R}^{S\times A}$ that rates the association of a given label $a\in A$ with a certain input $s\in S$. The classification rule $\pi_C\in A^S$ simply selects (one of) the label(s) that achieves the highest score for the given inputs:
\begin{equation}
  \label{eq:greedy2}
\pi_C(s) \in \argmax_a q(s,a).
\end{equation}
For example, \emph{Multi-class Support Vector Machines}~\cite{guermeur2011generic} can be seen as SFMC$^2$ algorithms, the same can be said of the structured margin approach \cite{taskar2005learning} which we consider in the experimental setting. Other algorithms may be envisionned (see Subsec.~\ref{subsec:instanciation}).

Given a dataset $D_C = \{(s_i,a_i=\pi^E(s_i))_i\}$ of actions $a_i$ (deterministically) chosen by the expert on states $s_i$, we train such a classifier. The classification policy $\pi_C$ is not the end product we are looking for (that would be mere imitation of the expert, not IRL). What is of particular interest to us is the score function $q$ itself. One can easily notice the similarity between Eq.~\ref{eq:greedy2} and Eq.~\ref{eq:greedy} that describes the relation between the optimal policy in a MDP and its optimal action-value function. It is only natural to view the score function $q$ of the classifier as some kind of optimal action-value function for the classifier policy $\pi_C$. By inversing the Bellman equation Eq.~\ref{eq:bellman1} with $q$ in lieu of $Q^\pi_R$, one gets $R^C$, the reward function relative to our score/action-value function $q$ :
\begin{equation}
  \label{eq:rc}
  R^C(s,a) =q(s,a) - \sum_{\mathrm{s'}}P(s'|s,a)q(s',\pi_C(s')).
\end{equation}
As we wish to solve the general, or approximate, IRL problem where the transition probabilities $P$ are unknown, our reward function $R^C$ will be approximated with the help of information gathered by interacting with the system. We assume that another dataset $D_R = \{(s_j,a_j,s'_j)_j\}$ is available where $s'_j$ is the state an agent taking action $a_j$ in state $s_j$ transitionned to. Action $a_j$ needs not be chosen by any particular policy. The dataset $D_R$ brings us information about the dynamics of the system. From it, we construct datapoints
\begin{equation}
  \label{eq:rj}
  \{\hat r_j = q(s_j,a_j) - \gamma q(s'_j,\pi_C(s'_j))\}_j.
\end{equation}
As $s'_j$ is sampled accordingly to $P(\cdot|s_j,a_j)$ the constructed datapoints help building a good approximation of $R^C(s=s_j,a=a_j)$. A regressor (a simple least-square approximator can do but other solutions such as Gaussian Processes \cite{rasmussen2006gaussian} could also be envisionned) is then fed the datapoints $((s_j,a_i),\hat r_j)$ to obtain $\hat R^C$, a generalisation of $\{((s_j,a_j),\hat r_j)_j\}$ over the whole state-action space. The complete algorithm is given in Alg.~\ref{algo:cascading}.
\begin{algorithm}%[H]
    %\small
  \caption{CSI algorithm}
  \label{algo:cascading}
  \emph{\textbf{Given}} a training set $D_C=\{(s_i,a_i=\pi^E(s_i))\}_{1\leq i \leq D}$ and another training set $D_R=\{(s_{j},a_{j},s'_{j})\}_{1\leq j \leq D'}$\;\\
  \emph{\textbf{Train}} a score function-based classifier on $D_C$, obtaining decision rule $\pi_C$ and score function $q:S\times A \rightarrow \mathbb R$\;\\
  \emph{\textbf{Learn}} a reward function $\hat R^C$ from the dataset $\{((s_{j},a_{j}),\hat{r}_j)\}_{1\leq j \leq D'}$, $\forall (s_j,a_j,s'_j) \in D_R,\hat{r}_j=q(s_{j},a_{j})-\gamma q(s'_{j},\pi_C(s'_{j}))$\;\\
  \emph{\textbf{Output}} the reward function $\hat R^{C}$ \;
\end{algorithm}

Cascading two supervised approaches like we do is a way to inject the MDP structure in the resolution of the problem. Indeed, mere classification only takes into account information from the expert (i.e., which action goes with which state) whereas using the Bellman equation in the expression of $\hat r_j$ makes use of the information lying in the transitions $(s_j,a_j,s'_j)$, namely information about the transition probabilities $P$. The final regression step is just a way to generalize this information about $P$ to the whole state-action space in order to have a well-behaved reward function.

Being able to alleviate the ill effects of scalability or data scarcity by leveraging the wide range of techniques developped for the classification and regression problems is a strong advantage of the CSI approach.
